The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-04-12 paper.tex
correct new structure

\section{Ontological Framework}
%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.

In our case, the ANNO-core (\cref{fig:core}) functions as a TO.
It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
\enquote{Spatial relator} (for describing measurements between landmarks, e.g., distances or angles),
\enquote{Phenotype} (to model properties that can be derived from the measurements) and \enquote{Anatomical spatial entity} (to describe anatomical directions, axes and planes).

The subclasses of the ANNO-Core classes build a DO, the ANNO-domain.
These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
specific landmarks on them (e.g., mentale dexter and mentale sinister),
concrete measurements/relators (e.g., the distance between mentale dexter and mentale sinister),
individual phenotypes (e.g., sex or height) and defined anatomical directions (e.g., cranial or basal),
axes (e.g., longitudinal or sagittal axis) and planes (e.g., frontal or sagittal plane).

@@

One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class).

@@

\section{Description of ANNO-core}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

@@

\section{Development of ANNO-domain}

\begin{figure}[h!t]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}

@@

\section{Use Case: Integration into \aw{}}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{aw3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:

>>>>

\section{Ontological Architecture}\label{sec:architecture}
%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNO-core, ANNO-domain, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild

%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.

In our case, the ANNO-core, shown in \cref{fig:core}, functions as a TO, see \cref{sec:core}.
It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
\enquote{Spatial relator} (for describing measurements between landmarks, e.g., distances or angles),
\enquote{Phenotype} (to model properties that can be derived from the measurements) and \enquote{Anatomical spatial entity} (to describe anatomical directions, axes and planes).

The subclasses of the ANNO-Core classes build a DO, the ANNO-domain, see \cref{sec:domain}.
These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
specific landmarks on them (e.g., mentale dexter and mentale sinister),
concrete measurements/relators (e.g., the distance between mentale dexter and mentale sinister),
individual phenotypes (e.g., sex or height) and defined anatomical directions (e.g., cranial or basal),
axes (e.g., longitudinal or sagittal axis) and planes (e.g., frontal or sagittal plane).

@@

One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class), as described in \cref{sec:aw}.

@@

\section{Description of ANNO-core}\label{sec:core}
% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

@@

\section{Development of ANNO-domain}\label{sec:domain}
% Arbeit der Domänenexperten, Tabelle, OWL-Generator
\begin{figure}[h!t]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}

@@

\section{Use Case: Integration into \aw{}}\label{sec:aw}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{aw3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:
